---
title: "Using the ScPCAr package to access Single-cell Pediatric Cancer Atlas data"
output: html_document
vignette: >
  %\VignetteIndexEntry{Using the ScPCAr package to access Single-cell Pediatric Cancer Atlas data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts}
#| include: FALSE
knitr::opts_chunk$set(
  fig.width = 6,
  fig.height = 4,
  dpi = 300,
  fig.path = "scpcar-"
)
```


# Introduction

The ScPCAr package provides an interface to interact with the Single-cell Pediatric Cancer Atlas (ScPCA) Portal API.
This vignette demonstrates the basic workflow for discovering, downloading, and working with ScPCA data.

This vignette covers the following tasks for interacting with the ScPCA Portal:

1. Listing available projects
2. Selecting a project and exploring its samples
3. Obtaining an authentication token
4. Downloading all data for an ScPCA project
5. Creating a custom dataset with specified samples and data types
6. Loading ScPCA data into R

# Installing the ScPCAr Package

The ScPCAr package is currently available via GitHub.
You can install the latest version using the `pak` package:

```{r install}
#| eval: FALSE
# Install pak if needed
if (!requireNamespace("pak", quietly = TRUE)) {
  install.packages("pak")
}

# Install ScPCAr from GitHub
pak::pak("Alexslemonade/ScPCAr")
```

You can then load the `ScPCAr` package.
Note that in this vignette we will use `ScPCAr::` when calling functions from the package, both for clarity and to avoid any possible namespace conflicts.
We will also load the `SingleCellExperiment` package for later analysis.

```{r setup}
#| message: FALSE
library(ScPCAr)
library(SingleCellExperiment)
```

# Exploring available projects

## Listing all projects

First, let's see what projects are available in the ScPCA Portal:

```{r list-projects}
# Get a data frame of all projects
projects <- ScPCAr::scpca_projects()

# print out a portion of the data frame
head(projects)
```

The `scpca_projects()` function returns a data frame with basic project metadata.
By default, it returns a simplified version with list columns removed for easier viewing.
You can see the full structure with additional data such as a list of diagnoses, data types, external accession numbers, etc., by setting `simplify = FALSE`:

```{r list-projects-full}
# Get the full project information including list columns
projects_full <- ScPCAr::scpca_projects(simplify = FALSE)

# View the structure of the full data frame
dplyr::glimpse(projects_full)
```


## Getting detailed project information

Now let's get more detailed information about the samples in a selected project.
We will use the first project, `SCPCP000001`, as an example.
According to the project info above, these samples are from a study of pediatric high-grade gliomas.
We can get more detailed information about this project using its `project_id`, and here we will set `simplifyVector = TRUE` to convert simple lists into vectors where possible.

```{r project-info}
project_id <- "SCPCP000001"

# Get detailed project metadata
project_info <- ScPCAr::get_project_info(project_id)
```

This returns a `list` with more detailed information about the project and the samples within.
You can explore the full structure of this list with `str()` or `dplyr::glimpse()`, but for now we will just look at a few of the components that might be of interest.

For example, we can look at the set of diagnoses that are present, and their counts:

```{r project-diagnoses}
data.frame(count = unlist(project_info$diagnoses_counts)) |>
  tibble::rownames_to_column("diagnosis")
```

Similarly, we can look at the data modalities available for this project, which here includes both single-cell and bulk RNA-seq data:

```{r project-modalities}
project_info$modalities
```


# Exploring sample information

## Getting sample metadata

Let's look at the samples within our selected project:

```{r sample-info}
# Get sample information for the project
sample_info <- ScPCAr::get_project_samples(project_id)
head(sample_info)
```

## Getting detailed sample information

We can also get detailed information for a specific sample:

```{r detailed-sample}
sample_id <- "SCPCS000001"
# Get detailed sample metadata
sample_detail <- ScPCAr::get_sample_info(sample_id)
str(sample_detail, max.level = 1)

# Check which data modalities are available for this sample
sample_detail$modalities
```

# Authentication

To download full data files, or to create a custom dataset, you must agree to the terms of service and obtain an authentication token.
To view the terms of service in a web browser, run `ScPCAr::view_terms()`.
Then you can use the `ScPCAr::get_auth()` function, providing your email address and agreeing to the terms of service:

```{r auth}
#| eval: FALSE
# Get an authentication token
# Replace with your actual email address
my_email <- "your.email@example.com"
auth_token <- ScPCAr::get_auth(email = my_email, agree = TRUE)
```

In addition to saving your authorization token to `auth_token` (or whatever variable name you set), the `ScPCAr::get_auth()` function will store the token in the environment variable `SCPCA_AUTH_TOKEN` so it can be accessed by other `ScPCAr` functions automatically.
If you have a saved token from a previous session, you can populate the environment with a command like: `Sys.setenv("SCPCA_AUTH_TOKEN" = auth_token)` or by setting the environment variable in your shell before launching `R`.

**Important Notes:**

- You must set `agree = TRUE` to indicate you accept the terms of service
- Replace `"your.email@example.com"` with your actual email address
- Keep your token secure and don't share it publicly
- If you create a custom dataset using `ScPCAr::create_dataset()`, you will need to use the same token to access that dataset later, so be sure to save it somewhere safe (e.g., a password manager).
  - In a later session, you can use `Sys.setenv(SCPCA_AUTH_TOKEN = saved_token)` to set the environment variable with your saved token.

# Downloading data for an ScPCA project

There are two main ways for downloading data from the ScPCA Portal using `ScPCAr`: 

- Project downloads contain all available samples for a given project
- Custom datasets allow you to specify specific subsets of samples, including samples from separate projects

We will start with project downloads first, as they are a bit simpler, though less flexible. 

## Downloading a project SingleCellExperiment objects

We will use the `download_project()` function, specifying the project ID, desired destination directory, and the file format.
We also have the option to download a merged object with all samples contained in a single object.
The function will download and unpack the files associated with that sample, and return a list of file paths for the downloaded files

Let's start with SingleCellExperiment format:

```{r download-project-sce}
#| message: FALSE
# Download SingleCellExperiment data for our project
# This will create a directory structure under "scpca_data/"
file_paths <- ScPCAr::download_project(
  project_id,
  destination = "scpca_data",
  format = "sce"
)

# List some of the files that were downloaded
head(file_paths)
```

### Project download file structure

A standard download of SingleCellExperiment data for a project will create a directory structure like this:

```
scpca_data
└── {project_id}_single-cell-experiment_{YYYY-MM-DD}
    ├── README.md
    ├── {project_id}_bulk_rna
    │   ├── {project_id}_bulk_metadata.tsv
    │   └── {project_id}_bulk_quant.tsv
    └── {project_id}_single-cell
        ├── {sample1_id}
        │   ├── {library1_id}_celltype-report.html
        │   ├── {library1_id}_filtered.rds
        │   ├── {library1_id}_processed.rds
        │   ├── {library1_id}_qc.html
        │   └── {library1_id}_unfiltered.rds
        ├── {sample2_id}
        ...
        
```

Note that the `library_id` and `sample_id` often have different numbers, as a single sample may have been sequenced across multiple libraries.
For details about the contents of these files, see the [ScPCA Portal documentation](https://scpca.readthedocs.io/en/latest/download_files.html).

## Downloading projects in AnnData/H5AD format

`ScPCAr` also supports downloading data in AnnData (H5AD) format, which is commonly used in Python-based single-cell analysis workflows:

```{r h5ad-discussion}
#| eval: FALSE
# Download the same project in H5AD format
file_paths_h5ad <- ScPCAr::download_project(
  project_id,
  destination = "scpca_data",
  format = "anndata"
)
```

For more about the expected files when downloading samples in this format, see the [ScPCA Portal AnnData docs](https://scpca.readthedocs.io/en/stable/download_files.html#anndata-downloads).


# Custom ScPCA datasets

The ScPCAr package supports the creation of custom datasets that allow you to choose specific samples for download, including samples from multiple projects.
After exploring the metadata with functions like `get_project_info()`, `get_project_samples()` and `get_sample_info()`, you may have a list of samples that you would like to download. 

## Creating a custom dataset

To create a custom dataset, use the `create_dataset()` function.
This will return a dataset ID that you will want to keep track of for later access to the dataset.
Note that the dataset is also tied to the authentication token, so you will need to use the same token to access the dataset later. 
Make sure you have it saved somewhere safe!

```{r}

```




# Working with ScPCA data in R

## Loading data into R


```{r load-sce}
# select the processed SCE files using a pattern match for the file name.
processed_sce_files <- stringr::str_subset(file_paths, "_processed\\.rds$")

# Load the first SingleCellExperiment object
sce <- readRDS(processed_sce_files[1])

# View a summary of the object
sce
```

### Working with the SingleCellExperiment object

Once loaded, you can work with the SingleCellExperiment object using standard Bioconductor tools.
For more information on the contents of the SingleCellExperiment objects provided by ScPCA, see the [ScPCA Portal documentation](https://scpca.readthedocs.io/en/latest/sce_file_contents.html#components-of-a-singlecellexperiment-object).

```{r work-with-sce}
# Access library metadata
library_metadata <- metadata(sce)
library_metadata$sample_id

# Access count data
counts <- counts(sce)
dim(counts)

# Access cell metadata
cell_metadata <- colData(sce)
cell_metadata

# Access gene metadata
gene_metadata <- rowData(sce)
gene_metadata
```

And of course we can make some standard plots, here a UMAP colored by the consensus cell type annotation.
(This one has a few too many cell types to be easily distinguishable, but we'll do it anyway for demonstration purposes only!)

```{r plot-umap}
#| fig.width: 10
#| fig.height: 6
#| fig.cap: "UMAP of sample SCPCS000001 colored by consensus cell type annotation."
# View the UMAP, colored by consensus cell type
scater::plotUMAP(sce, color_by = "consensus_celltype_annotation") +
  ggplot2::coord_fixed() +
  ggplot2::theme_classic() +
  ggplot2::guides(color = ggplot2::guide_legend(title = "Cell type")) +
  # remove axis ticks
  ggplot2::theme(
    axis.ticks = ggplot2::element_blank(),
    axis.text = ggplot2::element_blank()
  )
```



# Session info

<details>
<summary>Click to expand R session info</summary>
```{r session-info}
sessioninfo::session_info()
```
</details>
